---
title: "Exploring PCA Analysis - and how it works"
output: html_document
---


```{r}
expr_matrix_file = "test.matrix" #Trinity_trans.TMM.EXPR.matrix"
data = read.table(expr_matrix_file)
log2data = log2(data+1)
```

Scale the expression data in the columns according to Z-scores and center at zero.

```{r}
scaled_log2data = scale(log2data, center=T, scale=T)
```

Now center the expression for each gene

```{r}
pca_data = t(scale(t(scaled_log2data), center=T, scale=F))
```

Consider remvoing those genes that have low variance

```{r}
trans_sd = apply(pca_data, 1, sd)
hist(trans_sd, br=30)
```

# Running PCA

## Running PCA using svd

First, transpose the matrix so that the genes (the observations) are the columns and the samples are the rows.

```{r}
pca_data = t(pca_data)
```

Now, following how the 'prcomp' works under the hood:
https://svn.r-project.org/R/trunk/src/library/stats/R/prcomp.R

```{r}
s = svd(pca_data, nu=0)
sdev = s$d / sqrt(max(1, nrow(pca_data) - 1))
rotation = s$v   # eigenvalues or weights or loadings
pc_scores =  pca_data %*% rotation

head(rotation)
head(pc_scores)
plot(pc_scores[,1], pc_scores[,2])
```

Of course, you could just run the 'prcomp' like so:

```{r}
pca = prcomp(pca_data)
rotation = pca$rotation
pc_scores = pca$x

head(rotation)
head(pc_scores)
plot(pc_scores[,1], pc_scores[,2])
```

## Running PCA using Eigen

Another way to do this the longer way is to use eigen to extract the eigenvalues and eigenvectors from a correlation matrix

```{r}
trans_cov_matrix = cov(pca_data)
trans_cov_matrix[1:5,1:5] # just for show
myEig = eigen(trans_cov_matrix)
names(myEig)
#head(myEig$vectors)
scores = pca_data %*% myEig$vectors
scores[,1:15]
plot(scores[,1], scores[,2])
```